---
title: "CUSTARD User's Guide"
author: "Kamil Slowikowski"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set(
  cache = TRUE,
  autodep = TRUE,
  echo = FALSE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  cache.path = "cache/",
  fig.path = "figures/",
  concordance = TRUE
)

library(custard)
library(ggplot2)
library(cowplot)
library(reshape2)
library(scales)
library(RColorBrewer)

running_time <- function(n = 50, magnitude = 2, timepoints = 5) {
  sapply(2:timepoints, function(i) {
    n ^ 2 * (2 * magnitude + 1) ^ (i - 1)
  })
}

plot_template <- function(mat, cus, i = "23", text_size = 6) {
  d1 <- data.frame(
    Timepoints = rownames(cus$templates),
    Template = cus$templates[, i],
    stringsAsFactors = FALSE
  )
  d1$Timepoints <- as.integer(d1$Timepoints)
  
  idx <- cus$matches[i, ] > 0
  
  scale_factor <- abs(
    Reduce("-", quantile(mat[5,idx], probs = c(0.025, 0.975)))
  )
  d1$Template <- d1$Template / max(scale_factor, 1 / scale_factor)
  
  d2 <- melt(mat[, idx], varnames = c("Timepoint", "Gene"))
  
  ggplot() +
    geom_line(
      data = d2,
      aes(Timepoint, value, group = Gene),
      color = "blue", alpha = 0.25
    ) +
    geom_line(
      data = d1,
      aes(Timepoints, Template, group = 1),
      size = 2
    ) +
    annotate(
      geom = "text",
      x = -Inf,
      y = Inf,
      vjust = 1.1,
      hjust = -0.1,
      size = text_size,
      label = sprintf(
        "Template %s\n%s genes\nP = %s",
        i,
        scales::comma(sum(idx)),
        signif(cus$results[i, "BinomP"], 1)
      )
    ) +
    scale_x_discrete(breaks = rownames(mat), labels = rownames(mat)) +
    labs(x = "Timepoint", y = "Log ratio")
}

plot_templates <- function(selected_templates, cus = NULL) {
  timepoints <- nrow(selected_templates)
  d <- as.data.frame(t(selected_templates))
  d$Index <- 1:nrow(d)
  d <- melt(d, id.vars = c("Index"))
  d$Time <- rep(1:timepoints, each = ncol(selected_templates))
  
  ggplot(d, aes(x = Time, y = value)) +
    facet_wrap(~ Index) +
    geom_line() +
    scale_y_continuous(limits = c(-max(abs(d$value)), max(abs(d$value)))) +
    scale_x_discrete(breaks = pretty_breaks()) +
    labs(x = "", y = "")
}
```

# Introduction

CUSTARD finds pre-defined expression profiles (templates) that are overabundant
in short time course gene expression data.

The algorithm consists of the following steps:

1. Create a small set of expression profile templates.
2. Assign genes to each template.
3. Test each template for enrichment with gene expression profiles.

CUSTARD is an R implementation of the work by Jason Ernst et al. described in:

> Ernst, J., Nau, G. J. & Bar-Joseph, Z. Clustering short time series gene expression data. Bioinformatics 21 Suppl 1, i159–68 (2005). https://www.ncbi.nlm.nih.gov/pubmed/15961453

We have extended the method to meet the specific needs of our analysis, and we
provide those extensions in the CUSTARD R package.

This document is meant to help new users understand the concepts underlying the
method.

# Templates

We want to find overabundant expression profiles in a short time series gene
expression dataset. To do this, we will assume a simple model and create a few
profiles that will serve as templates.

After each time point, we expect that a gene will do one of the following:

- stay the same
- increase in expression
- decrease in expression

In our model, the gene will increase or decrease in discrete units. To better
understand what the landscape of such a model looks like, let's visualize all
possible templates with `n = 5` timepoints and allowing `c = 1, 2, 3` units of
change at each time point:

```{r all_templates_lines, echo=FALSE, fig.height=3, fig.width=8}
plot_all_templates <- function(timepoints = 5, magnitude = 1) {
  all_templates <- make_templates(timepoints = timepoints, magnitude = magnitude)
  d <- as.data.frame(t(all_templates))
  d$Index <- 1:nrow(d)
  d <- melt(d, id.vars = c("Index"))
  d$Time <- rep(1:timepoints, each = ncol(all_templates))
  ggplot(d, aes(x = Time, y = value, color = factor(Index), group = Index)) +
    geom_line() +
    theme(legend.position = "none") +
    # scale_y_continuous(breaks = seq(min(d$value), max(d$value))) +
    scale_y_continuous(breaks = pretty_breaks()) +
    labs(
      x = "Time point", y = "Log ratio", title = sprintf("c = %d", magnitude)
    )
}

cowplot::plot_grid(
  plot_all_templates(magnitude = 1),
  plot_all_templates(magnitude = 2),
  plot_all_templates(magnitude = 3),
  ncol = 3
)
```

At time point 2, notice that the templates under `c = 1` are allowed to stay
the same as time point 1, increase by one unit, or decrease by one unit. The
gene has 3 choices at each transition between time points, for a total of
$3^4 = 81$ distinct templates. 

We can also visualize the 81 templates for `c = 1` as a heatmap:

```{r all_templates_heatmap, fig.height=2, fig.width=8}
all_templates <- make_templates(timepoints = 5, magnitude = 1)
par(mar=c(0,0,0,0))
image(
  x = t(all_templates)[,nrow(all_templates):1],
  xaxt = "n",
  yaxt = "n",
  frame.plot = FALSE,
  col = brewer.pal(11, "Spectral")
)
```

Use a greedy algorithm to select `m = 10` distinct profiles from the 81
possibilities when `c = 1`:

```{r distinct_templates, fig.height=6, fig.width=8}
timepoints <- 5
all_templates <- make_templates(timepoints = timepoints, magnitude = 1)
selected_templates <- distinct_templates(all_templates, templates = 10)

plot_templates(selected_templates)
```

Running time of template selection algorithm:

```{r running_time}
maxt <- 12
d <- melt(data.frame(
  Timepoints = 2:maxt,
  "c = 1" = running_time(n = 50, magnitude = 1, timepoints = maxt),
  "c = 2" = running_time(n = 50, magnitude = 2, timepoints = maxt),
  "c = 3" = running_time(n = 50, magnitude = 3, timepoints = maxt),
  check.names = FALSE
), id.vars = "Timepoints")

ggplot(d, aes(x = Timepoints, y = value, group = variable, color = variable)) +
  geom_line() +
  geom_text(
    data = subset(d, Timepoints == max(d$Timepoints)),
    aes(label = variable),
    nudge_y = 0.1, nudge_x = 0.5,
    size = 5
  ) +
  # scale_x_continuous(breaks = 2:8) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
  # scale_y_log10() +
  theme(legend.position = "none") +
  labs(
    x = "Number of time points",
    y = "Total running time",
    title = "Running time to generate m = 50 distinct templates"
  )
```

# Enrichment

Test enrichment with 5,000 null genes sampled from the uniform distribution
against 25 templates with `c = 1`:

```{r test_enrichment_code, eval=FALSE, echo=TRUE}
mat <- matrix(runif(5000 * 5), nrow = 5)
cus <- custard(mat, templates = 25, magnitude = 1)
```

```{r test_enrichment}
set.seed(42)
mat <- matrix(runif(5000 * 5), nrow = 5)
colnames(mat) <- paste("G", 1:5000, sep = "")
rownames(mat) <- 1:5
mat <- log10(t(apply(mat, 1, function(x) x / mat[1, ])))

# set.seed(42)
cus <- custard(mat, templates = 25, magnitude = 1)

# The results are the same for unscaled or scaled templates, because
# we use the Pearson correlation coefficient to match genes to templates.

# set.seed(42)
# all_templates <- make_templates(timepoints = nrow(mat), magnitude = 1)
# selected_templates <- distinct_templates(all_templates, templates = 25)
# # Scale the templates to match the range of 95% of the data.
# mat_range <- abs(
#   Reduce("-", quantile(mat[nrow(mat),], probs = c(0.025, 0.975)))
# )
# selected_templates <- selected_templates / mat_range
# 
# cus2 <- custard(mat, templates = selected_templates)
# 
# cus$results <- cus$results[order(rownames(cus$results)), ]
# cus2$results <- cus2$results[order(rownames(cus2$results)), ]

# plot(-log10(cus$results$BinomP), -log10(cus2$results$BinomP)); abline(0, 1)
# plot(-log10(cus$results$PermP), -log10(cus2$results$PermP)); abline(0, 1)

ggplot() +
  geom_point(
    data = cus$results,
    aes(Expected, Observed)
  ) +
  geom_line(
    data = cus$binom_qvals,
    aes(Expected, BinomQ)
  )

# ggplot(data = cus$results, aes(-log10(PermP), -log10(BinomP))) +
#   geom_point()
```

Here is a table of the top 6 templates:

```{r enrichment_test_table}
kable(head(cus$results))
```

Let's take a closer look at template 23, the one with the strongest P value:

```{r enrichment_test_template}
plot_template(mat, cus, "8", text_size = 5) +
  theme_cowplot(font_size = 16)
```
